Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2024-08-08 paper/main.tex
Updates

Figure shows some of these issues when applied to the Adult Income dataset. Due
to these challenges, even when using automated algorithms, we often have to
rely on manual post hoc testing and modifications to the model learned by the
algorithm. Whether constructing the model manually or using an algorithm, we
must rely on domain expertise to construct or modify the model. Hence, it is
crucial to have tools to guide this modification process. One of the ways to
guide this modification is to use Conditional Independence (CI) based model
testing \cite{Ankan2023}. As DAGs imply CI statements that can be tested in the
given data using statistical CI tests, we can verify whether the implied CIs
are consistent in the data or not. If some of implied CIs are not valid in the
given dataset, we can modify the model accordingly. Figure~\ref{fig:ci_testing}
shows an example of model testing using CIs. However, using p-values of the CI
tests has a few drawbacks. Firstly, as CIs are implied only by missing edges in
the model, we can not test whether an incorrect edge is already present in the
model. For examples, in high sample size scenarios, we might get a significant
association between a pair of variables with very small association. Secondly,
the number of implied CIs can be quite large. For example, in the moderately
sized alarm network with $ 37 $ node, there are $287$ implied CIs. If we rely
only on the p-values of the CI tests, it can be difficult to figure which test
to focus on.

To deal with both this issues, in this paper, we develop a measure of partial
association which can be using in combination of the CI test to rank the
violations in these CI tests. On the other hand, in Structural Equation Models
(SEMs) literature, more emphasis has been placed on developing tools to help
researchers manually construct models while integrating their domain expertise.

@@

			\draw[->]  (age) to[bend right=20] (ed);
			\draw[->]  (sex) to (ed);
			\draw[->]  (age) to (hrpw);
			\draw[->]  (ed) to (hrpw);
			\draw[->]  (sex) to (hrpw);
			\draw[->]  (ed) to (income);
			\draw[->]  (hrpw) to (income);
			\draw[->]  (occ) to[out=300, in=0, looseness=1.4] (income.east);
			\draw[->]  (sex) to (income);
			\draw[->]  (ed) to[out=90, in=150, looseness=1.3] (occ);
			\draw[->]  (sex) to (occ);	
		\end{scope}
		\begin{scope}[xshift=-3cm]
			\draw[thick, ->, double, cyan] (2.5,0) -- (3,0);
			\node[rectangle, align=center, inner sep=1pt] at (5, 0) {
				\begin{minipage}{.2\textwidth}
					\begin{equation*}
						\begin{split}
							\textnormal{Age} &\ci \textnormal{Incm} \rvert \textnormal{Edct, Hrpw, Sex} \\
							\textnormal{Age} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
							\textnormal{Age} &\ci \textnormal{Sex} \\
							\textnormal{Hrpw} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
						\end{split}
					\end{equation*}
				\end{minipage}
				};
		\end{scope}
		\end{tikzpicture}
		\caption{\todo[inline]{Finish figure. Show an example of model learnt using PC algorithm $\rightarrow$ Orienting the unoriented edges $\rightarrow$  Model testing}}
		\label{fig:ci_testing}
\end{figure}

A common method for specification search is using \emph{modification indices},
which identify potential changes to the model structure that improve the model
fit. These potential changes are then ranked by the ones that improve the fit
the most. This ranking allows the researchers to focus first on the suggestions
that can improve the model the most and while integrating their expertise to
choose the most optimal modifications. Inspired by this ranking approach in
modification indices, in this paper we develop a method to rank the CI tests
using a conditional measure of association. Our proposed approach uses the
measure of association to find correlated variables in the data that are not
explained by the current model. Using the measure, we can then rank the pair of
variables between which adding an edge would improve the explanation the most.
Based on this information, the user can select a potential pair of variables
and based on their expert knowledge decide the direction of the edge.

Our main contributions in this paper are as follows:
\begin{enumerate}
	\item We present a measure of (conditional) association for mixed data
		based on canonical correlations. This measure of association
		generalizes the widely used partial correlation to mixed data
		setting.
	\item Using this partial association measure, we present a procedure to
		rank CI tests. This ranking can be used to prioritize addition
		or deletion of edges from the model.
	\item Lastly, we present a both a web-browser based tool and an
		implementation in the Python package to allow researchers to
		apply this method to their own datasets.
\end{enumerate}

@@

In this section, we develop a novel partial measure of association for mixed
data by extending the idea of partial correlation to mixed data. As partial
correlations require two things: 1) A residualization method, and 2) measure of
associations between the residuals. We generalize this to mixed data by combining a
mixed data residualization method \citep{Ankan2023} with a multivariate measure
of association based on Pillai's Trace. Given $ n $ samples $ D = (x, y,
\bm{z}) $ of variables $ X $, $ Y $, and $ \bm{Z} $, we are interested in
computing an estimate of the conditional association $ \rho(X, Y; \bm{Z}) $
between $ X $ and $ Y $ conditioned on $ \bm{Z} $. As the residualization
approach requires atleast one conditional variable to be present, if $ \bm{Z} =
\emptyset $, we use a constant value for $ Z $. Similar to partial correlation,
we first compute the residuals for $ X $ and $ Y $ using $ \bm{Z} $ as the
predictor variables. We use the mixed data residualization approach from
\citet{Ankan2023}. Depending on the type of dependent variable, the residuals
are defined as follows:

@@

\section{Connection to score based methods}

\begin{itemize}
	\item Because they make only local changes, can get stuck in local minimas.
\end{itemize}

>>>>

Figure~\ref{fig:intro} shows some of these issues when applied to the Adult
Income dataset. Due to these challenges, whether constructing the model
manually or using an algorithm, we typically have to rely on some expert
knowledge to get a model that we can use for further analysis. Hence, it is
crucial to have tools to guide this construction or modification process.

A common method for specification search is using \emph{modification indices},
which identify potential changes to the model structure that improve the model
fit. These potential changes are then ranked by the ones that improve the fit
the most. This ranking allows the researchers to focus first on the suggestions
that can improve the model the most and while integrating their expertise to
choose the most optimal modifications.

One of the ways to guide this modification is to use Conditional Independence
(CI) based model testing \cite{Ankan2023}. As DAGs imply CI statements that can
be tested in the given data using statistical CI tests, we can verify whether
the implied CIs are consistent with the data or not. If some of implied CIs are
not valid in the given dataset, we can modify the model accordingly.
Figure~\ref{fig:ci_testing} shows an example of model testing using CIs.
However, using p-values of the CI tests has a few drawbacks. Firstly, as CIs
are implied only by missing edges in the model, we can not test whether an
incorrect edge is already present in the model. For examples, in high sample
size scenarios, we might get a significant association between a pair of
variables with very small association. Secondly, the number of implied CIs can
be quite large. For example, in the moderately sized alarm network with $ 37 $
node, there are $287$ implied CIs. If we rely only on the p-values of the CI
tests, it can be difficult to figure which test to focus on.

Inspired by this ranking approach in modification indices, in this paper we
develop a method to rank the CI tests using a conditional measure of
association.  To deal with both this issues, in this paper, we develop a
measure of partial association which can be using in combination of the CI test
to rank the violations in these CI tests. On the other hand, in Structural
Equation Models (SEMs) literature, more emphasis has been placed on developing
tools to help researchers manually construct models while integrating their
domain expertise. Our proposed approach uses the measure of association to find
correlated variables in the data that are not explained by the current model.
Using the measure, we can then rank the pair of variables between which adding
an edge would improve the explanation the most. Based on this information, the
user can select a potential pair of variables and based on their expert
knowledge decide the direction of the edge.

@@

			\draw[->]  (age) to[bend right=20] (ed);
			\draw[->]  (sex) to (ed);
			\draw[->]  (age) to (hrpw);
			\draw[->]  (ed) to (hrpw);
			\draw[->]  (sex) to (hrpw);
			\draw[->]  (ed) to (income);
			\draw[->]  (hrpw) to (income);
			\draw[->]  (occ) to[out=300, in=0, looseness=1.4] (income.east);
			\draw[->]  (sex) to (income);
			\draw[->]  (ed) to[out=90, in=150, looseness=1.3] (occ);
			\draw[->]  (sex) to (occ);	
		\end{scope}
		\begin{scope}[xshift=-3cm]
			\draw[thick, ->, double, cyan] (2.5,0) -- (3,0);
			\node[rectangle, align=center, inner sep=1pt] at (5, 0) {
				\begin{minipage}{.2\textwidth}
					\begin{equation*}
						\begin{split}
							\textnormal{Age} &\ci \textnormal{Incm} \rvert \textnormal{Edct, Hrpw, Sex} \\
							\textnormal{Age} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
							\textnormal{Age} &\ci \textnormal{Sex} \\
							\textnormal{Hrpw} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
						\end{split}
					\end{equation*}
				\end{minipage}
				};
		\end{scope}
		\end{tikzpicture}
		\caption{\todo[inline]{Finish figure. Show an example of model learnt using PC algorithm $\rightarrow$ Orienting the unoriented edges $\rightarrow$  Model testing}}
		\label{fig:intro}
\end{figure}

Our main contributions in this paper are as follows:
\begin{enumerate}
	\item We present a measure of (conditional) association for mixed data
		based on canonical correlations. This measure of association
		generalizes the widely used partial correlation to mixed data
		setting.
	\item Using this partial association measure, we present a procedure to
		rank CI tests. This ranking can be used to prioritize addition
		or deletion of edges from the model.
	\item Lastly, we present both a web-browser based tool and an
		implementation in the Python package to allow researchers to
		apply this method to their own datasets.
\end{enumerate}

@@

In this section, we develop a novel partial measure of association for mixed
data by extending partial correlations to mixed data. As partial correlations
require two things: 1) A residualization method, and 2) measure of associations
between the residuals. We generalize this to mixed data by combining a mixed
data residualization method \citep{Ankan2023} with a multivariate measure of
association based on Pillai's Trace. Given $ n $ samples $ D = (x, y, \bm{z}) $
of variables $ X $, $ Y $, and $ \bm{Z} $, we are interested in computing an
estimate of the conditional association $ \rho(X, Y; \bm{Z}) $ between $ X $
and $ Y $ conditioned on $ \bm{Z} $. As the residualization approach requires
atleast one conditional variable to be present, if $ \bm{Z} = \emptyset $, we
use a constant value for $ Z $. Similar to partial correlation, we first
compute the residuals for $ X $ and $ Y $ using $ \bm{Z} $ as the predictor
variables. We use the mixed data residualization approach from
\citet{Ankan2023}. Depending on the type of dependent variable, the residuals
are defined as follows:

@@

\section{Connection to Score Based Methods}
\begin{itemize}
	\item The process of iteratively adding and removing edges in very similar to
		Greedy Equivalence Search (GES).
	\item But Because they make only local changes, can get stuck in local minimas.
	\item This approach looks for modifications globally.
\end{itemize}

\todo[inline]{Show an example where local changes can get stuck}
